Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood

### 2025-09-19 paper/supplemental_information.Rmd
work on implementing SI changes

---
title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
  - \renewcommand{\thefigure}{S\arabic{figure}}
  - \renewcommand{\thetable}{S\arabic{table}}
---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
library(here)
source(here("helper","common.R"))
```

```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE}
paper <- suppressMessages(knitr::purl(here("paper","paper.Rmd"), documentation = 0))
suppressMessages(source(paper))
```

@@

```{r, eval=F}
#run Creel prep script

#creel prep script expects files in root directory
setwd(here("supplemental_data","creel_2024"))
source("Step2aCorrsNDataPrep.R")
setwd(here())
# run Peekbank-style rt computation over Creel data
source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))

 creel_plot <- ggplot(filter(subj_rt_acc,n_correct_trials>1&n_trials>1),aes(mean_rt,mean_rt_correct))+
  geom_point(aes(size=n_correct_trials),alpha=0.3,fill="black",stroke=NA)+
  geom_smooth(method="lm")+
  xlab("Mean Participant Reaction Time (ms)\nAll Trials")+
  ylab("Mean Participant Reaction Time (ms)\nTrials w/ Correct Pointing Response")+
  theme_cowplot(font_size=20)+
  #add correlation and p-value as text box
  annotate("text",x=500,y=1500,label=paste0("r = ",round(cor(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs"),2)),size=9)+
  theme(legend.position=c(0.8,0.2))+
  scale_size_continuous(name="# Trials",breaks=c(2,5,10,15,20))
ggsave(here("supplemental_data","creel_2024","creel_rt_correlation_correcttrials_vs_alltrials.png"),width=9,height=6)

@@

```{r}
# <!--To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024). This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image. Using this dataset, we investigated the correlation between reaction times (following the same procedure as in our main analyses, i.e. focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent. The results are visualized below. Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`). This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children. While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.-->

```{r, eval=F}
creel_plot
```

@@

@@

BIC(norm)
BIC(lnorm)
BIC(wald)
BIC(exgaus)

rt_fits <- rt_fitting |>
  group_by(age_bin) |>
  nest(rt) |>
  mutate(
    norm = map(data, \(d) fitdist(d$rt, distr = "norm", method = "mle")),
    lnorm = map(data, \(d) fitdist(d$rt, distr = "lnorm", method = "mle")),
    wald = map(data, \(d) fitdist(d$rt, distr = "invgauss", method = "mle", 
                                  start = list(mean = 1, shape = 1, dispersion = 1))),
    exgaus = map(data, \(d) fitdist(d$rt, distr = "exGAUS", method = "mle", 
                                    start = list(mu = 1000, sigma = 500, nu = 1), 
                                    lower = c(0, 0, 0)))) |>
  select(-data) |>
  pivot_longer(norm:exgaus, names_to = "distribution", values_to = "fit") |>
  mutate(BIC = map_dbl(fit, BIC)) |>
  group_by(age_bin) |>
  mutate(min_bic = BIC == min(BIC))

ggplot(rt_fits, 
       aes(x = distribution, y = BIC, col = min_bic)) + 
  geom_point(stat = "identity", position = "dodge") + 
  scale_color_solarized(name = "Lowest BIC") + 
  xlab("Age bin") + 
  ylab("BIC") + 
  coord_flip() + 
  facet_wrap(~age_bin, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r}
rt_hist_data <- d_sub |>
  mutate(
    age_bin = cut(age,
                  breaks = c(0,12,15,18,21,24,27,30,36,60), 
                  labels = c("<12", "12-15","15-18", "18-21", 
                             "21-24", "24-27","27-30","30-36", ">36")))

@@

```{r}
acc_hist_data <- d_sub |>
  mutate(
    age_bin = cut(age,
                  breaks = c(0,12,15,18,21,24,27,30,36,60), 
                  labels = c("<12", "12-15","15-18", "18-21", 
                             "21-24", "24-27","27-30","30-36", ">36")))

@@

norm <- fitdist(acc_fitting$long_window_accuracy, distr = "norm", method = "mle")
beta <- fitdist(acc_fitting$long_window_accuracy, distr = "beta", method = "mle")

BIC(norm)
BIC(beta)
```

acc_fits <- acc_fitting |>
  group_by(age_bin) |>
  nest(long_window_accuracy) |>
  mutate(
    norm = map(data, \(d) fitdist(d$long_window_accuracy, distr = "norm", method = "mle")),
    beta = map(data, \(d) fitdist(d$long_window_accuracy, distr = "beta", method = "mle"))) |>
  select(-data) |>
  pivot_longer(norm:beta, names_to = "distribution", values_to = "fit") |>
  mutate(BIC = map_dbl(fit, BIC)) |>
  group_by(age_bin) |>
  mutate(min_bic = BIC == min(BIC))

@@

\verbatim{lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)}

@@

```{r}
#remotes::install_github("dr-JT/semoutput")
library(semoutput)
semoutput::sem_fitmeasures(fit3)
semoutput::sem_fitmeasures(fit3_firstadmin)
semoutput::sem_fitmeasures(fit3_age)
semoutput::sem_fitmeasures(fit3_age_firstadmin)

@@

summary(mod_nlme)

@@

#random effects
dataset_preds <- d_sub |>
  filter(!is.na(prod)) |> # filter on predictor
  group_by(dataset_name) |>
  summarise(min_age = min(age),
            max_age = max(age)) |>
  group_by(dataset_name) |>
  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0_resid = c(quantile(d_sub_prod$log_rt_0_resid, 
                                                                                                     c(.25,.75), 
                                                                                                     na.rm=TRUE))))) |>
  select(-min_age, -max_age) |>
  unnest(col = "df") |>
  ungroup()

@@

```{r}
library(brms)

@@

# Fit the model
mod_brms <- brm(
  formula = nlform,
  data = d_sub_prod,
  prior = priors,
  family = gaussian(),
  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  # backend = "cmdstanr",  # optional but faster
  save_pars = save_pars(all = TRUE),
  sample_prior = "yes"
)

@@

@@

# Fit the model
resid_mod_brms <- brm(
  formula = nlform_resid,
  data = d_sub_prod,
  prior = priors_resid,
  family = gaussian(),
  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  # backend = "cmdstanr",  # optional but faster
  save_pars = save_pars(all = TRUE),
  sample_prior = "yes"
)

@@

@@

ggplot(new_data_resid, aes(x = age_c, y = fit, col = as.factor(resid_log_rt_0_c))) +
  geom_line(linewidth = 2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(resid_log_rt_0_c), group = as.factor(resid_log_rt_0_c)), 
              alpha = 0.2) +
  labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
  theme_minimal() + 
  scale_color_viridis() + 
  geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black")

@@

```{r}
colMeans(!is.na(d_sub_wide_fine))
```

>>>>

---
title: "Supplemental Information: Continuous developmental changes in word recognition support language learning across early childhood"
output: 
  pdf_document:
    fig_caption: yes
    keep_tex: yes
header-includes:
  - \usepackage{threeparttablex}
  - \usepackage{booktabs}
  - \renewcommand{\thefigure}{S\arabic{figure}}
  - \renewcommand{\thetable}{S\arabic{table}}
  - \usepackage{float}
  - \renewcommand{\textfraction}{0.00}
  - \renewcommand{\topfraction}{1}
  - \renewcommand{\bottomfraction}{1}
  - \renewcommand{\floatpagefraction}{1}
  - \setcounter{topnumber}{5}
  - \setcounter{bottomnumber}{5}
  - \setcounter{totalnumber}{6}

---

```{r, echo=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo=FALSE, message=FALSE, warning=FALSE, cache=TRUE)
library(here)
source(here("helper","common.R"))
library(papaja)
```

```{r, echo=FALSE, cache=TRUE, message=FALSE, warning=FALSE, include=F}
paper <- suppressMessages(knitr::purl(here("paper","paper.Rmd"), documentation = 0))
suppressMessages(source(paper))
```

@@

```{r, include=F}
#run Creel prep script

#creel prep script expects files in root directory
detach("package:dplyr", character.only = TRUE)
library("dplyr", character.only = TRUE)
setwd(here("supplemental_data","creel_2024"))
source("Step2aCorrsNDataPrep.R")
setwd(here())
# run Peekbank-style rt computation over Creel data
source(here("supplemental_data","creel_2024","creel_peekbank_acc_rt_analysis.R"))

@@

To address this concern, we investigated data from a recent large-scale word recognition study with toddlers in which eyetracking measures were collected together with a final pointing response (Creel, 2024). This dataset included 914 responses from children (2.5-6.5 years) completing a looking-while listening procedure in which they also were instructed to point to the target image. Using this dataset, we investigated the correlation between reaction times (following the same procedure as in our main analyses, i.e. focusing specifically on distractor-to-target shifts) computed over all trials and reaction times computed only for those trials in which children selected the correct referent. The results are shown in Figure \ref{fig:creelplot}. Reaction times (i.e., distractor to target shifts) for correct trials only were highly correlated with reaction times across all trials (`r apa_print(creel_correlation)$full_result`). This result suggests that having the ability to filter out incorrect trials has a minimal impact on reaction time computation, even in young children. While there is some uncertainty about how these results may generalize to infants in our younger age ranges (i.e., below 2.5 years of age), who struggle to provide reliable pointing responses, it seems reasonable to assume that our reaction time results would stay largely the same if it were possible to filter out trials on which infants make an incorrect mapping between the target label and the target image using an eyetracking-independent final choice response.

```{r creel-plot, fig.cap="\\label{fig:creelplot} Correlation between reaction times on all trials and reaction times on trials where the child pointed to the correct target. Data from Creel (2024).", fig.pos="h!", fig.width=8, fig.height=6}

ggplot(filter(subj_rt_acc,n_correct_trials>1&n_trials>1),aes(mean_rt,mean_rt_correct))+
  geom_point(aes(size=n_correct_trials),alpha=0.3,fill="black",stroke=NA)+
  geom_smooth(method="lm")+
  xlab("Mean Participant Reaction Time (ms)\nAll Trials")+
  ylab("Mean Participant Reaction Time (ms)\nTrials w/ Correct Pointing Response")+
  theme_cowplot(font_size=20)+
  #add correlation and p-value as text box
  annotate("text",x=500,y=1500,label=paste0("r = ",round(cor(filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt,filter(subj_rt_acc,n_correct_trials>1&n_trials>1)$mean_rt_correct,use="pairwise.complete.obs"),2)),size=7)+
  theme(legend.position=c(0.8,0.2))+
  scale_size_continuous(name="# Trials",breaks=c(2,5,10,15,20))
```

@@

(TODO someone should edit this text)
A potential concern with multiple measures deriving from the same underlying data (as with RT/accuracy in the looking-while-listening paradigm) is that distributional assumptions about each variable could implicitly assume some connections between RT/accuracy and variability that might not be warranted.

To begin, we note that RT/accuracy and variability could be connected due to true age-related covariation - that is, some common cause leading to improvement in both, namely improvement in speed of response and improvement in consistency of response (e.g., skill learning). But they could be necessarily connected by virtue of their distributional form, and this is the concern we address here.

@@

#BIC(norm)
#BIC(lnorm)
#BIC(wald)
#BIC(exgaus)

rt_fits <- rt_fitting |>
  group_by(age_bin) |>
  nest(rt) |>
  mutate(
    norm = map(data, \(d) fitdist(d$rt, distr = "norm", method = "mle")),
    lnorm = map(data, \(d) fitdist(d$rt, distr = "lnorm", method = "mle")),
    wald = map(data, \(d) fitdist(d$rt, distr = "invgauss", method = "mle", 
                                  start = list(mean = 1, shape = 1, dispersion = 1))),
    exgaus = map(data, \(d) fitdist(d$rt, distr = "exGAUS", method = "mle", 
                                    start = list(mu = 1000, sigma = 500, nu = 1), 
                                    lower = c(0, 0, 0)))) |>
  select(-data) |>
  pivot_longer(norm:exgaus, names_to = "distribution", values_to = "fit") |>
  mutate(BIC = map_dbl(fit, BIC)) |>
  group_by(age_bin) |>
  mutate(min_bic = BIC == min(BIC))
```

We’ll look at reaction time first. Under certain distributional forms for reaction time, there would be a necessary relationship between RT and its variation. However, the literature focuses on the use of the Exponential-Gaussian (ex-Gaussian) distribution as well as Wald, Weibull, gamma, and log normal distributions (see for example Luce, 1986; Ratcliff, 1993; Van Zandt, 2002). All of these are 2- or 3-parameter distributions meaning that there is no necessary relationship between mean and variance.

The problem of fitting RT distributions is complex and a substantial literature exists (e.g., Ratcliff, 1979; Luce, 1986; Van Zandt, 2000; Baayen & Milin, 2010). One of the big challenges in our dataset as well as elsewhere is that distributions are conditional on factors such as participant and task, so it is challenging to draw inferences about the underlying distribution when looking at average data.

That said, we have attempted some fitting and we find that overall the data are best fit by either an ex-Gaussian or a log normal distribution, again consistent with prior literature, giving us confidence in this conclusion. Across the full dataset, the BIC values for ex-Gaussian (`r BIC(exgaus)`) and log normal (`r BIC(lnorm)`) are quite close to one another, and better than the Wald (`r BIC(wald)`) and normal (`r BIC(norm)`) fits. When binned by age (Fig \ref{fig:rt-fits}), younger children seem better fit for a log normal distribution and older children seem better fit by ex-Gaussian (models with lowest BICs are shown in red since significant differences can be obscured by the large scale). Figure \ref{fig:rt-hist} shows the RT data distribution overlaid with the corresponding log-normal distributions.

Overall, we think this result generally vitiates our decision to use log-transformed RTs as our primary dependent measure, and further suggests that reaction time means and variabilities are, in principle, theoretically independent.

```{r rt-fits, fig.cap="\\label{fig:rt-fits} Goodness of fit for different distributional models for RT, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
ggplot(rt_fits, 
       aes(x = distribution, y = BIC, col = min_bic)) + 
  geom_point(stat = "identity", position = "dodge") + 
  scale_color_solarized(name = "Lowest BIC") + 
  xlab("Age bin") + 
  ylab("BIC") + 
  coord_flip() + 
  facet_wrap(~age_bin, scales = "free_y") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

```{r rt-hist, fig.cap="\\label{fig:rt-hist} Distribution of RT overlaid with log normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=6}

rt_hist_data <- d_sub |>
  mutate(
    age_bin = cut(age,
                  breaks = c(0,12,15,18,21,24,27,30,36,60), 
                  labels = c("<12", "12-15","15-18", "18-21", 
                             "21-24", "24-27","27-30","30-36", ">36")))

@@

norm_acc <- fitdist(acc_fitting$long_window_accuracy, distr = "norm", method = "mle")
beta_acc <- fitdist(acc_fitting$long_window_accuracy, distr = "beta", method = "mle")

#BIC(norm)
#BIC(beta)
```

Individual trial-level accuracies are not binomial because they are an average probability of fixation over a viewing window, as described in the manuscript. They are bounded at 0 and 1, but in general they tend towards the range .5 - .8 in most studies of this population. Figure \ref{acc-hist} shows the data binned by age group with fitted gaussian distributions.

These distributions seem well fit by standard gaussians, but they are in principle bounded and so we asked whether this made a difference, using a Beta distribution (a two-parameter continuous distribution bounded at 0 and 1) for fitting. Surprisingly, across all data, the BIC values for the two distributions were very similar (`r BIC(norm_acc)` for normal versus `r BIC(beta_acc)` for Beta), though the Beta was slightly favored. Across age groups, there was heterogeneity with some groups better fit by a gaussian and others better fit by a Beta (Figure \ref{fig:acc-fits}).

Again, we feel that this result generally vitiates our approach of modeling accuracies via standard linear mixed effects models: their distributional form is quite close to normal. Thus, again there is no necessary connection between mean and variance: they could in principle be independent. As above, we will plan on including this information in the Supplement.

```{r acc-hist, fig.cap="\\label{fig:acc-hist} Goodness of fit for different distributional models of accuracy, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
acc_hist_data <- d_sub |>
  mutate(
    age_bin = cut(age,
                  breaks = c(0,12,15,18,21,24,27,30,36,60), 
                  labels = c("<12", "12-15","15-18", "18-21", 
                             "21-24", "24-27","27-30","30-36", ">36")))

@@

```{r acc-fits, fig.cap="\\label{fig:acc-fits} Distribution of accuracies overlaid with normal distribution, split by age.", fig.pos="h!", fig.width=8, fig.height=6}
acc_fits <- acc_fitting |>
  group_by(age_bin) |>
  nest(long_window_accuracy) |>
  mutate(
    norm = map(data, \(d) fitdist(d$long_window_accuracy, distr = "norm", method = "mle")),
    beta = map(data, \(d) fitdist(d$long_window_accuracy, distr = "beta", method = "mle"))) |>
  select(-data) |>
  pivot_longer(norm:beta, names_to = "distribution", values_to = "fit") |>
  mutate(BIC = map_dbl(fit, BIC)) |>
  group_by(age_bin) |>
  mutate(min_bic = BIC == min(BIC))

@@

<!--\verbatim{lmer(long_window_acc_var ~ log_age_s + (log_age_s | dataset_name) + (1 | subject_id)}-->

FIXME because I cause things to crash!

@@

```{r}
#remotes::install_github("dr-JT/semoutput")
library(semoutput)
semoutput::sem_fitmeasures(fit3, print=F) |> kable()
semoutput::sem_fitmeasures(fit3_firstadmin, print=F) |> kable()
semoutput::sem_fitmeasures(fit3_age, print=F) |> kable()
semoutput::sem_fitmeasures(fit3_age_firstadmin, print=F) |> kable()

@@

#summary(mod_nlme)

@@

#random effects
dataset_preds <- d_sub |>
  filter(!is.na(prod)) |> # filter on predictor
  group_by(dataset_name) |>
  summarise(min_age = min(age),
            max_age = max(age)) |>
  group_by(dataset_name) |>
  mutate(df = map2(min_age, max_age, ~expand_grid(age = seq(.x, .y, .1), log_rt_0_resid = c(quantile(d_sub_prod$log_rt_0_resid, 
                                                                                                     c(.25,.75), 
                                                                                                     na.rm=TRUE))))) |>
  dplyr::select(-min_age, -max_age) |>
  unnest(col = "df") |>
  ungroup()

@@

```{r, eval=F}
library(brms)

@@

# Fit the model
mod_brms <- brm(
  formula = nlform,
  data = d_sub_prod,
  prior = priors,
  family = gaussian(),
  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  # backend = "cmdstanr",  # optional but faster
  save_pars = save_pars(all = TRUE),
  sample_prior = "yes",
  file="brms1.rds"
)

@@

saveRDS(fitted_vals, "brms1_fitted.rds")
```

```{r}
fitted_vals <- readRDS("brms1_fitted.rds")

@@

# Fit the model
resid_mod_brms <- brm(
  formula = nlform_resid,
  data = d_sub_prod,
  prior = priors_resid,
  family = gaussian(),
  # init = 0,  # initialize all parameters to 0 on unconstrained scale (safe for centered priors)
  chains = 4, cores = 4,
  control = list(adapt_delta = 0.95),
  # backend = "cmdstanr",  # optional but faster
  save_pars = save_pars(all = TRUE),
  sample_prior = "yes",
  file="brms2.rds"
)

@@

saveRDS(fitted_vals, "brms2_fitted.rds")
```

fitted_vals <- readRDS("brms2_fitted.rds")

@@

ggplot(new_data_resid, aes(x = age_c, y = fit, col = as.factor(resid_log_rt_0_c))) +
  geom_line(linewidth = 2) +
  geom_ribbon(aes(ymin = lower, ymax = upper, fill = as.factor(resid_log_rt_0_c), group = as.factor(resid_log_rt_0_c)), 
              alpha = 0.2) +
  labs(x = "Centered Age", y = "Predicted prod (fixed effects only)") +
  theme_minimal() + 
  scale_color_viridis(discrete=T) + 
  scale_fill_viridis(discrete=T, guide=F)+
  geom_point(data = d_sub_prod, aes(x = age_c, y = prod), col = "black")

@@

```{r, eval=F}
colMeans(!is.na(d_sub_wide_fine))
```
